LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-06-14 article/main.tex
final corrections

\textbf{Large language models (LLMs)} have become a commonplace research topic
today.
Because language models such as BERT~\cite{devlin2018bert} or
GPT~\cite{gpt-2018,gpt2-2019,gpt3-2020} constantly advance the state of the art
on many natural language processing tasks, it is interesting to evaluate them on
more specialized tasks.
Multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} show that large
language models are good at following instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

@@

Taking Noam Chomsky's ideas~\cite{chomsky2002syntactic} that syntax and meaning
are different things, we intuit that writing style is defined in terms of both.
We shy away from discussing meaning, but syntax and orderly constructs like it
seem to be within our grasp.
Traditionally, linguistic style revolved around morphological and syntactic
arrangements~\cite{lugea2023stylistics}.
More recent descriptions find semantics and pragmatics to be inextricably linked
to author style~\cite{verma2019lexical}, but argue that a great deal can still
be inferred from lexicon, syntax and a `surface' level which contains basic HLFs
such as the number of words in a sentence.
Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Linguistic style is the sum of identifiable language choices manifest in a text
made from the language system by the text producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
Opinions that style depends on context and author choice vis-a-vis a
communication goal seem to converge~\cite{mcdonald1985computational,
    hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
Our selection of HLFs is informed by aspects of linguistic style.

@@

Our approach stands out for its simplicity and focus.
We propose engineering instruction prompts for LLMs, a straightforward strategy
distinct from complex alternatives.
With discussions on style transfer or author verification beyond this paper's
scope, we present a method controlling LLM text generation through the use of
HLFs.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
LFTK~\cite{lftk-2023} is a framework built on top of spaCy~\cite{spacy}
that provides implementations for computing many HLFs.
We use these two libraries to perform our experiment.
They categorize HLFs by domain and family and make available information on the
suitability of each HLF for given tasks such as readability assessment or fake
news detection.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

@@

Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
instruction result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
We define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$.

@@

In the next sections we present our perspective on the most relevant results of
the experiment.

@@

Highly abstract HLF instructions are unlikely to have the desired effect.
Furthermore, the impact tends to be adverse.
This suggests that LLMs are particularly adept at following instructions
regarding HLFs when those instructions are more concrete.

@@

Other experiments are required to better understand the capabilities of large
language models with regard to handcrafted linguistic features.
The selection of language models, the reliance on a corpus for computing target
linguistic features, the fixed choice of input text are all limitations of the
current approach.
Using a selection of HLFs and not individually analysing the behaviour of the
LLM under the influence of each individual HLF is another limitation of our
approach.

@@

We began by trying to understand whether it is possible to generate text that
exhibits certain linguistic features by instructing a large language model.
To our surprise, it turns out that instructing large language models to write
text that conforms to a selected assortment of linguistic features leads to
significantly different linguistic featues of the generated text.
This is especially true for concrete instructions.

However, the outcomes are not always desirable.
From a pragmatic standpoint, prompt engineering and a careful choice of language
features and input text seem like a good idea to alleviate results.
Providing examples in the input prompt does not seem as beneficial.

\bibliography{references}
\newpage
\appendix

>>>>

\textbf{Large language models (LLMs)} have become a commonplace research topic
today.
Because language models such as BERT~\cite{devlin2018bert} or
GPT~\cite{gpt-2018,gpt2-2019,gpt3-2020} constantly advance the state of the art
on many natural language processing tasks, it is interesting to evaluate them on
more specialized tasks.
Multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} show that large
language models are good at following instructions.
We intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

@@

Taking Noam Chomsky's ideas that syntax and meaning are different
things~\cite{chomsky2002syntactic}, we intuit that writing style is defined in
terms of both.
We shy away from discussing meaning, but syntax and similarly orderly constructs
seem to be within our grasp.
Traditionally, linguistic style revolved around morphological and syntactic
arrangements~\cite{lugea2023stylistics}.
More recent descriptions find semantics and pragmatics to be inextricably linked
to author style~\cite{verma2019lexical}, but argue that a great deal can still
be inferred from lexicon, syntax and a `surface' level which contains basic HLFs
such as the number of words in a sentence.
Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Linguistic style is the sum of identifiable language choices manifest in a text
made from the language system by the text producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
Opinions seem to converge that style depends on context and author choice
vis-a-vis a communication goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
Our selection of HLFs is informed by aspects of linguistic style.

@@

Our approach stands out through its simplicity and focus.
We propose engineering instruction prompts for LLMs, a straightforward strategy
distinct from complex alternatives.
With discussions on style transfer or author verification beyond this paper's
scope, we present a method controlling LLM text generation through the use of
HLFs.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
LFTK~\cite{lftk-2023} is a framework built on top of spaCy~\cite{spacy}
that provides implementations for computing many HLFs.
We use these two libraries to perform our experiment.
LFTK categorizes HLFs by domain and family and make available information on the
suitability of each HLF for given tasks such as readability assessment or fake
news detection.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

@@

Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
We define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$.

@@

Next, we present our perspective on the most relevant results of the experiment.

@@

Highly abstract HLF instructions are unlikely to have the desired effect.
Furthermore, the impact tends to be adverse.
This suggests that LLMs are more adept at following instructions regarding HLFs
when those instructions are more concrete.

@@

Other experiments are required to better understand the capabilities of large
language models with regard to HLFs.
The selection of language models, the reliance on a corpus for computing target
linguistic features, the fixed choice of input text are all limitations of the
current approach.
Making different experiment design choices in all these respects might shed more
light on this matter.
Using a selection of HLFs and not individually analysing the behaviour of the
LLM under the influence of each individual HLF is another limiting choice that
invites to future work on the alternative.

@@

We began by trying to understand whether it is possible to generate text that
exhibits certain linguistic features by instructing a large language model.
To our surprise, it turns out state of the art large language models are
receptive to instructions regarding the linguistic features of the output.
This is especially true for concrete instructions.

However, the outcomes are not always good.
From a pragmatic standpoint, prompt engineering and a careful choice of language
features and input text seem like a good idea to alleviate results.
Providing examples in the input prompt does not seem as beneficial.

\newpage

\bibliography{references}

\newpage

\appendix
